---
title: 'Vectorisation, `purrr` and `furrr`'
author:
- name: "Stephen Ashton"
  email: "stephen.ashton@dhsc.gov.uk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

The following packages are required for the examples below. Feel free to install
as required:

```{r, eval=FALSE}
install.packages("rlang")
install.packages("dplyr")
install.packages("purrr")
install.packages("future")
install.packages("furrr")
```

# Linear Operations

When working with simple pieces of data and analysis in R, it is often fine to
apply modifications in-line and sequentially.

For example, we may wish to multiply two numbers together:

```{r}
prod <- 3 * 6

prod
```

# Loops

As we work using more complex code, we may wish to apply a modification to a
larger amount of data - likely contained in vectors or lists.

A simple way to do this would be to work with a `for`-loop. For example, if we
wish to work out the square of a vector of numbers, we can do this as follows:

```{r}
# Specify input
numbers <- 1:10

numbers

# Initial output
squares <- c()

# Run for loop
for (i in seq_along(numbers)) {
  squares[i] <- numbers[i] ^ 2
}

squares
```

However, an issue with such a process will be that the computation time will
extend with the length of the input vector and complexity of the function
applied. The above uses a small vector and simple function so is very quick, 
but with a much longer loop or more complex function may result in a much slower
result.

# Vectorisation

The easiest solution to fix this is to use **vectorisation**. This applies a
function to all elements of a vector (or set of vectors) simultaneously, and is
highly efficient within R.

For example, we can do the same process as the previous example as a vectorised
implementation:

```{r}
# Specify input
numbers <- 1:10

numbers

# Apply vectorisation
squares <- numbers ^ 2

squares
```

We can also apply a calculation using two different vectors (of equal length),
for example:

```{r}
# Specify inputs
numbers <- 1:10
powers <- seq(from = 2, to = 20, by = 2)

numbers
powers

# Apply vectorisation
big_nums <- numbers ^ powers

big_nums
```

However, an issue with this is that the function being applied must either be
an in-built vectorised function, or be constructed in such a way that it can be
applied in this way. Additionally, the vector(s) must be **atomic** - that is
that every item in the vector must be of the same length and type (e.g. they
must all be numerics (as in the above) or strings).

# Mapping with `purrr`

An efficient way to apply the same function to a list/vector of values (or
multiple lists/vectors) is to use
the `purrr` package, notably the functions:

*   `purrr::map`
*   `purrr::map2`
*   `purrr::pmap`

With these we can deal with more complex scenarios and functions. For example,
we may have a list containing vectors, and wish to calculate the mean of each
entry:

```{r}
# Load library
suppressWarnings(suppressPackageStartupMessages(library(purrr)))

# Specify input
val_list <- list(
  c(1,2,3),
  c(1,1,2,2,3,3,4,4),
  c(2,3,4,4,6)
)

val_list

# Apply map
mean_list <- map(val_list, mean)

mean_list
```

It should be noted that (by default) `purrr::map` will always return a list. If
we want a specific type, we can use the `purrr::map_*` family of functions, or
apply post-processing. For example, in the above example, if we want to return
a numeric vector, we could do the following:

```{r}
# Apply map
mean_list <- map_dbl(val_list, mean)

mean_list
```

## How to use `purrr::map`

Within `purrr::map`, we need to specify (at least) two values:

*   `.x` - The list or vector
*   `.f` - A function

In the above example, we used an inbuilt function (`mean`) where the input was
clear, however this is often not the case. In such a scenario, we should use an
**anonymous function** to specify directly. For example:

```{r}
# Specify input
val_list <- list(
  c(1,2,3),
  c(1,1,2,2,3,3,4,4),
  c(2,3,4,4,6)
)

val_list

# Apply map
meansq_list <- map_dbl(
  .x = val_list, 
  .f = \(x) mean(x, na.rm = TRUE) ^ 2
)

meansq_list
```

You can also pass additional arguments to `.f` within the map - for example, the
following two statements are equivalent:

```{r}
mean_list <- map_dbl(val_list, mean, na.rm = TRUE)
mean_list <- map_dbl(val_list, \(x) mean(x, na.rm = TRUE))
```

The first version should be avoided except where absolutely necessary as it can
become confusing and/or unclear.

### Note on Anonymous Functions

If you have not come across anonymous functions before, these are very simple to
construct - you specify the inputs to the function with a leading slash, then
specify the mechanism.

For example:

```{r}
f1 <- \(x) mean(x, na.rm = TRUE)
f2 <- \(para1, para2) para1 ^ para2
f3 <- \(a,b,c,d) (a + b) - (c ^ d) + 3
```

You can also nest these functions:

```{r}
f4 <- \(x, y) map(x, \(x_val) x_val + y)
```

If you have used anonymous functions before, you may be used to using the `~`
notation with `.x`, `.y` and `.z`, or `..1`, `..2` etc. as parameters. This
approach is no longer advised as it is unclear - for example, using this
approach, this is how `f4` would be written:

```{r}
f4 <- ~ map(.x, ~ .x + .y, .y = .y)
```

This is messy and will cause issues if you are using `lintr` and/or
`devtools::check()` to verify your code integrity. **Do not do this or you will
be cursed by Hadley!**

## Extending to `purrr::map2` and `purrr::pmap`

If you want to extend to two (or more) parameter functions instead, we can use
`purrr::map2` and `purrr::pmap`, and their relatives `purrr::map2_*` and
`purrr::pmap_*`.

For example, rewriting our earlier example:

```{r}
# Specify inputs
numbers <- 1:10
powers <- seq(from = 2, to = 20, by = 2)

numbers
powers

# Apply vectorisation
big_nums <- map2_dbl(numbers, powers, \(x,y) x ^ y)

big_nums
```

For `purrr::pmap` the input should be a named list of lists. The names of these
lists should be the same as the parameters in the anonymous function:

```{r}
# Load library for dataset
suppressWarnings(suppressPackageStartupMessages(library(dplyr)))

# Specify inputs
inputs_list <- list(
  character_nm <- starwars$name,
  character_hgt <- starwars$height,
  character_wgt <- starwars$mass,
  character_typ <- starwars$species
)

# Apply pmap
description <- pmap_chr(
  inputs_list,
  \(character_nm, character_hgt, character_wgt, character_typ) paste(
    character_nm,
    "is a",
    character_typ,
    "character in the Star Wars franchise. They are",
    character_hgt,
    "cm tall and weigh approximately",
    character_wgt,
    "kg under standard Earth gravity."
  )
)
  
description[1:5]
```

## Using `purrr::walk` instead of `purrr::map`

On occasion, we may wish to have the effects of a function, but not
necessarily care about the return. In these cases, we can replace `purrr::map`,
`purrr::map2` and `purrr::pmap` with `purrr::walk`, `purrr::walk2` and
`purrr::pwalk`.

For example, if we want to use `rlang::warn` to print a warning to console,
then most of the time we will not care about the actual return - in this case we
can use a `walk`-style function instead of a `map`-style function:

```{r}
# Load library
suppressWarnings(suppressPackageStartupMessages(library(rlang)))

# Apply walk to earlier output
walk(description[1:5], warn)
```

# Parallelisation using `furrr`

For applications of complex functions to large amounts of data, `purrr`-style
methods may still be slow. However, in these cases we can easily take advantage
of the resources available on modern computers to apply a several-fold
improvement in the runtime.

All of the techniques mentioned so far are **serial** functions - that is that
the use a single process running on the CPU at a given time. However, most
modern systems are actually constructed so that multiple processes can be run
concurrently - that is that we can have **parallel** processes.

```{r}
# Load libraries
suppressWarnings(suppressPackageStartupMessages(library(future)))
suppressWarnings(suppressPackageStartupMessages(library(furrr)))

availableCores()
```

Using the above command we can see that the latest generation of DHSC laptops
are built with `8` cores - which allows us to do (up to) `8` tasks concurrently.

## Considerations Regarding Parallel Computation

It should be noted that parallelisation isn't a "magic bullet" - there is some
overhead in terms of starting up and closing down the parallel process, so it
should only be used where the task is large enough to warrant this additional
time.

For example, if we assume that starting up and closing down the parallel
process takes 1s each then a task which takes 0.1s per cycle, and requires 1600
cycles then:

*   In serial, this would take 1600 * 0.1 = 160s
*   In parallel, this would take (1600/8) * (1.0 + 0.1 + 1.0) = 200 * 2.1 = 420s

However, if we assume the task takes 10s per cycle:

*   In serial, this would take 1600 * 10 = 16000s
*   In parallel, this would take (1600/8) * (1 + 10 + 1) = 200 * 12 = 2400s

It should be clear that the first task should be run in serial, whilst using a 
parallel approach for the second task is more efficient.

## Suitability for Parallel Computation

We should also note that not all calculations are suitable for use in a
parallel environment. For example, if we have a function such as the below:

```{r}
# Specify inputs
data <- 1:10

# Initialise outputs
outputs <- c()

# Run for loop
for (i in seq_along(data)) {
  outputs[i] <- sum(outputs) + data[i]
}

outputs
```

This function would _not_ be suitable for parallelisation as each entry is
dependent upon the previous result, which cannot be accessed using
parallelisation!

## Moving from `purrr` to `furrr`

As we already know how to write code in `purrr`, moving to `furrr` is (mostly)
very easy.

If we take our Star Wars example from above, we can do some simple conversion to
get the following:

```{r}
# Apply parallel map
description_parallel <- future_pmap_chr(
  inputs_list,
  \(character_nm, character_hgt, character_wgt, character_typ) paste(
    character_nm,
    "is a",
    character_typ,
    "character in the Star Wars franchise. They are",
    character_hgt,
    "cm tall and weigh approximately",
    character_wgt,
    "kg under standard Earth gravity."
  )
)
  
description_parallel[1:5]
```

In this example, we simply need to add `future_` in front of our `purrr`-based
function - it's that simple!

## Setting up the Environment

I lied. It isn't.

Well, it _almost_ is that simple, but the above code actually ran in serial, not
parallel, as we didn't specify our parallel specification.

This is done using the `future::plan()` function - there are four built-in
options for this:

*   `sequential` - this is the default plan and uses the `furrr` structure to
run in serial.
*   `multisession` - this is a parallel plan which uses the socketing version of
parallelisation on a single machine.
*   `multicore` - this is a parallel plan which uses the forking version of
parallelisation on a single machine. This can be more efficient than
`multisession`, but will **not** work on Windows-based computers.
*   `cluster` - the is a parallel plan which uses parallelisation across
multiple machines (for example, if you're working on a HPC system).

This plan is implemented as follows:

```{r}
# Setup parallel plan - specify plan and number of workers
plan(multisession, workers = 6)

# Apply parallel map
description_parallel <- future_pmap_chr(
  inputs_list,
  \(character_nm, character_hgt, character_wgt, character_typ) paste(
    character_nm,
    "is a",
    character_typ,
    "character in the Star Wars franchise. They are",
    character_hgt,
    "cm tall and weigh approximately",
    character_wgt,
    "kg under standard Earth gravity."
  )
)

# Restore default plan
plan(sequential)
  
description_parallel[1:5]
```

## Taking `furrr` Further

### Exporting Objects and Packages

If your `furrr` code uses external objects - such as packages or objects - then
these _should_ automatically be gathered and exported into the parallel jobs.
However, this does not always work - sometimes the computer gremlins don't like
you!

If this happens, don't worry - you can explicitly specify objects and packages
to export using `furrr::furrr_options()`. For example, say we want to explicitly
export the `inputs_list` object and the `dplyr` package to our jobs. This can be
done as follows:

```{r}
# Setup parallel plan - specify plan and number of workers
plan(multisession, workers = 6)

# Apply parallel map
description_parallel <- future_pmap_chr(
  inputs_list,
  \(character_nm, character_hgt, character_wgt, character_typ) paste(
    character_nm,
    "is a",
    character_typ,
    "character in the Star Wars franchise. They are",
    character_hgt,
    "cm tall and weigh approximately",
    character_wgt,
    "kg under standard Earth gravity."
  ),
  .options = furrr_options(
    globals = c("inputs_list"),
    packages = c("dplyr")
  )
)

# Restore default plan
plan(sequential)
  
description_parallel[1:5]
```

### Randomness in `furrr`

Working with random numbers in parallel is usually a nightmare - it is very
difficult to ensure consistency across machines and setups.

Thankfully, `furrr` makes this very easy! Unless explicitly disabled, randomness
within `furrr` is consistent, no matter the plan or number of workers! It will
return a different result to the same code in `purrr`, but is much better than
the alternative! For example:

```{r}
# Setup parallel plan - specify plan and number of workers
plan(multisession, workers = 6)

# Apply parallel map
rng_1 <- future_map(
  1:10,
  \(x) rnorm(x, mean = 0, sd = 1),
  # Specify master seed
  .options = furrr_options(seed = 1234)
)

# Setup parallel plan - specify plan and number of workers
plan(multisession, workers = 4)

# Apply parallel map
rng_2 <- future_map(
  1:10,
  \(x) rnorm(x, mean = 0, sd = 1),
  # Specify master seed
  .options = furrr_options(seed = 1234)
)

# Restore default plan
plan(sequential)

# Apply parallel map
rng_3 <- future_map(
  1:10,
  \(x) rnorm(x, mean = 0, sd = 1),
  # Specify master seed
  .options = furrr_options(seed = 1234)
)
```

As we can see the three results are identical:

```{r}
identical(rng_1, rng_2)
identical(rng_1, rng_3)
identical(rng_2, rng_3)
```